A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-15 latex/main.tex
Updates from Overleaf

The LS HMM at the heart of sc2ts provides a simple and powerful means 
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution for the 
putative recombinant sequence with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination  (similar to 
idea of ``parsimony improvement'' in RIPPLES.)
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

@@

Figure~\ref{fig:recombinant_evidence}A plots the number of mutations required
in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
and the resulting distribution of the number of 
averted mutations (lower histogram)
over all 354 recombination events that pass QC. 
Also shown is the 3SEQ classification status of these recombinants.
All of the 12\% of recombinants that do not pass 3SEQ validation are 
from the most weakly supported, with 4 or 5 averted mutations.
Figures~\ref{fig:recombinant_evidence_cov_recomb} and
~\ref{fig:recombinant_evidence_rebar} 
same the same information for CovRecomb and rebar validation,
showing that their validation status tends to be function 
of the divergence between the recombinant parents.

@@

To compare the accuracy of sc2ts with UShER+RIPPLES
we ran RIPPLES the Viridian UShER phylogeny
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
We used the RIPPLES parameter $p=4$ as is equivalent to the
sc2ts $k=4$ parameter used to build the ARG,
which resulted in 11[XX] distinct recombination events.
We then ran 3SEQ on each of the candidate trios,
giving an overall validation rate of XX\% (STAR Methods).
Comparing events identified in the UShER tree and sc2ts ARG
is not straightforward in general, and
so we focus comparison on the singleton recombinants
returned by both.
Although these are likely to be enriched for artefacts,
both methods face the same challenges and the 
events can be compared like-for like. Of the 11XX recombination
events returned by RIPPLES, 757 are singletons (resp 525/855 for 
sc2ts). Of the RIPPLES singletons, 705 are also singletons 
in the sc2ts ARG. [There's a complicated story to tell here 
about extra singletons in the RIPPLES set which pass
3SEQ validation, but have parents from the future.]
Thus, 216 of these recombinants pass 3SEQ validation, have 
at least one parent that is on average 63 days \emph{younger}
than the sequence itself. These parents were not available
to sc2ts as potential matches, as it enforces strict 
time ordering based on sample metadata.
Further study is required to determine
if the metadata is incorrect in these cases, or if these 
should be flagged as false positives. However, the 
key point is that RIPPLES is dependent on signals
embedded in the UShER tree, and therefore we cannot 
decouple the finer details of post-hoc recombinant detection
from the tree building algorithm.

We also compared the recombination events output by RIPPLES with
the Pango X lineages. At p=4, we found that only XX of the 
X lineages were present and many well-characterized events 
such as XBB were not present. We therefore reran at p=3,
resulting in XXXX events overall. At p=4, events involving
XX of the 44 Pango X lineages present in the dataset,
with unequivocal origins for 10 lineages, and 
7 lineages other lineages associated with two competing events. 
[More summary]

>>>>

The LS HMM at the heart of sc2ts provides a simple and powerful means
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution
for the putative recombinant sequence
with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination
(similar to the idea of ``parsimony improvement'' in RIPPLES.)
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

@@

Figure~\ref{fig:recombinant_evidence}A plots
the number of mutations required
in the recombinant versus non-recombinant
Viterbi solutions (upper scatter plot)
and the resulting distribution of the number of
averted mutations (lower histogram)
over all 354 recombination events that pass QC.
Also shown is the 3SEQ classification status of these recombinants.
All of the 12\% of recombinants that do not pass 3SEQ validation are
from the most weakly supported, with 4 or 5 averted mutations.
Figures~\ref{fig:recombinant_evidence_cov_recomb} and
~\ref{fig:recombinant_evidence_rebar} 
plot the same information for CovRecomb and rebar validation,
showing that their validation status tends to be function
of the divergence between the recombinant parents.

@@

To compare the accuracy of sc2ts with UShER+RIPPLES
we ran RIPPLES the Viridian UShER phylogeny
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
We used the RIPPLES parameter $p=4$ as is equivalent to the
sc2ts $k=4$ parameter used to build the ARG,
which resulted in 11[XX] distinct recombination events.
We then ran 3SEQ on each of the candidate trios,
giving an overall validation rate of XX\% (STAR Methods).
Comparing events identified in the UShER tree and sc2ts ARG
is not straightforward in general, and
so we focus comparison on the singleton recombinants
returned by both.
Although these are likely to be enriched for artefacts,
both methods face the same challenges and
the events can be compared like-for like.
Of the 11XX recombination events returned by RIPPLES,
757 are singletons (respectively, 525/855 for sc2ts).
Of the RIPPLES singletons, 705 are also singletons in the sc2ts ARG.
[There's a complicated story to tell here
about extra singletons in the RIPPLES set which pass
3SEQ validation, but have parents from the future.]
Thus, 216 of these recombinants pass 3SEQ validation,
have at least one parent that is on average
63 days \emph{younger} than the sequence itself.
These parents were not available to sc2ts as potential matches,
as it enforces strict time ordering based on sample metadata.
Further study is required to determine
if the metadata is incorrect in these cases, or
if these should be flagged as false positives.
However, the key point is that RIPPLES is dependent on signals
embedded in the UShER tree, and therefore we cannot
decouple the finer details of post-hoc recombinant detection
from the tree building algorithm.

We also compared the recombination events
output by RIPPLES with the Pango X lineages.
At p=4, we found that only XX of the Pango X lineages were present and
many well-characterized events such as XBB were not present.
We therefore reran at p=3, resulting in XXXX events overall.
At p=4, events involving XX of the 44 Pango X lineages
present in the dataset, with unequivocal origins for 10 lineages, and
7 lineages other lineages associated with two competing events.
[More summary]
